Supplementary information: Variational theory of non-relativistic quantum electrodynamics

### 2019-04-01 paper4pg.tex
moved discussion to supp, i think we're ok now.

The full Hamiltonian $H$, which depends on the fields $\psi$ and $\mathbf{A}$ is parameterized in terms of an orthonormal set of electron single-particle wavefunctions (orbitals) $\{\psi_n\}$, and in terms of a set of photonic mode functions (orbitals) $\{\mathbf{F}_i\}$. The electron field operator takes the form $\psi(\mathbf{x}) = \sum_n \psi_n(\mathbf{x})c_n$.
The $c_n$ is an annihilation operator for an electron corresponding to state $n$. The electromagnetic field operator takes the form $\mathbf{A}(\mathbf{x}) = \sum_i\sqrt{\frac{\hbar}{2\epsilon_0\omega_i}} \left(\mathbf{F}_i(\mathbf{x})a_i+\mathbf{F}^*_i(\mathbf{x})a^{\dagger}_i\right)$, where the $a_i^{(\dagger)}$ annihilate (create) a photon in mode $i$.The electromagnetic field operator is parameterized by both mode functions and  frequencies. The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}$ \cite{joannopoulos2011photonic}. \textcolor{blue}{We comment that in writing the Hamiltonian this way, we make a few assumptions. For one, that the interaction of matter with the quantized electromagnetic field is mediated solely by the vector potential. Moreover, we are assuming that the effect of free-space modes, which an emitter weakly interacts with, is taken into account by renormalization of the electron mass and charge \cite{milonni1976semiclassical}. We focus only on the interaction of the emitter with low-energy cavity modes where strong interaction is possible. For these cavity modes, a natural cutoff emerges based on the transparency of the cavity.}

@@

For the photon orbitals and energies, the minimization yields:
\begin{equation}
\left( \nabla\times\nabla\times - \frac{\omega_i^2}{c^2}\left(1-\frac{\omega_p^2(\mathbf{x})}{\omega_i^2} \right)\right)\mathbf{F}_i = 0,
\label{eq:mhf-photon}
\end{equation}
where $\omega_p^2(\mathbf{x}) = \frac{e^2}{m\epsilon_0}\sum\limits_{n=1}^N |\psi_n(\mathbf{x})|^2$ is a position-dependent squared-plasma frequency which will push the photon orbitals out of the region where the emitter is located. Equations (\ref{eq:mhf-electron}) and (\ref{eq:mhf-photon}) are main results and can be used to describe ultra-strongly coupled systems in three dimensions, in an arbitrary photonic system, and with multi-electron matter. Excited states in this framework can be identified with matter and photon quasiparticle excitations. \textcolor{blue}{Taking the divergence of Eq.~\ref{eq:mhf-photon}, we see that $\nabla\cdot(1-\frac{\omega_p^2(\mathbf{x})}{\omega^2})\mathbf{F}(\mathbf{x})=0$, which is a generalized Coulomb gauge condition on the modes. The generalized Coulomb gauge, $\nabla\cdot\epsilon(\mathbf{x})\mathbf{A}(\mathbf{x})=0$ for some dielectric permittivity $\epsilon(\mathbf{x}), is the most common gauge for analyzing phenomena in the discipline of macroscopic QED, in which one seeks to analyze the interaction of charges with fields in dielectric media \cite{glauber1991quantum}. The reason a particular gauge was chosen was that in the Hamiltonian, we have already assumed that all interaction with the quantized modes is manifested only through the vector potential, and that there is no scalar potential contribution to the interaction: relaxing this assumption allows more general gauge freedom. We also remark that if we performed a frame transformation, for example to the length gauge, the energy would be the same as that of our ansatz in the velocity gauge, although the state would be markedly different, and would be a highly correlated state.}

>>>>

The full Hamiltonian $H$, which depends on the fields $\psi$ and $\mathbf{A}$ is parameterized in terms of an orthonormal set of electron single-particle wavefunctions (orbitals) $\{\psi_n\}$, and in terms of a set of photonic mode functions (orbitals) $\{\mathbf{F}_i\}$. The electron field operator takes the form $\psi(\mathbf{x}) = \sum_n \psi_n(\mathbf{x})c_n$.
The $c_n$ is an annihilation operator for an electron corresponding to state $n$. The electromagnetic field operator takes the form $\mathbf{A}(\mathbf{x}) = \sum_i\sqrt{\frac{\hbar}{2\epsilon_0\omega_i}} \left(\mathbf{F}_i(\mathbf{x})a_i+\mathbf{F}^*_i(\mathbf{x})a^{\dagger}_i\right)$, where the $a_i^{(\dagger)}$ annihilate (create) a photon in mode $i$.The electromagnetic field operator is parameterized by both mode functions and  frequencies. The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}$ \cite{joannopoulos2011photonic}. \textcolor{blue}{Assumptions behind the form of the Hamiltonian are stated in Supplementary Materials, page 2.}

%\textcolor{blue}{In writing the Hamiltonian this way, we make a few assumptions. For one, that the interaction of matter with the quantized electromagnetic field is mediated solely by the vector potential. Moreover, we are assuming that the effect of free-space modes, which an emitter weakly interacts with, is taken into account by renormalization of the electron mass and charge \cite{milonni1976semiclassical}. We focus only on the interaction of the emitter with low-energy cavity modes where strong interaction is possible. For these cavity modes, a natural cutoff emerges based on the transparency of the cavity.}

@@

For the photon orbitals and energies, the minimization yields:
\begin{equation}
\left( \nabla\times\nabla\times - \frac{\omega_i^2}{c^2}\left(1-\frac{\omega_p^2(\mathbf{x})}{\omega_i^2} \right)\right)\mathbf{F}_i = 0,
\label{eq:mhf-photon}
\end{equation}
where $\omega_p^2(\mathbf{x}) = \frac{e^2}{m\epsilon_0}\sum\limits_{n=1}^N |\psi_n(\mathbf{x})|^2$ is a position-dependent squared-plasma frequency which will push the photon orbitals out of the region where the emitter is located. Equations (\ref{eq:mhf-electron}) and (\ref{eq:mhf-photon}) are main results and can be used to describe ultra-strongly coupled systems in three dimensions, in an arbitrary photonic system, and with multi-electron matter. Excited states in this framework can be identified with matter and photon quasiparticle excitations. \textcolor{blue}{Taking the divergence of Eq.~\ref{eq:mhf-photon}, we see that $\nabla\cdot(1-\frac{\omega_p^2(\mathbf{x})}{\omega^2})\mathbf{F}(\mathbf{x})=0$, which is a generalized Coulomb gauge condition on the modes \cite{glauber1991quantum}. For more discussion, see Supplementary Information, page 2.}

### 2019-04-01 supp2.tex
moved discussion to supp, i think we're ok now.

\maketitle
In this Supplement, we derive a self-consistent extension of the equations describing the ground state of a general quantum electrodynamical system with the effect of correlations included. We derive the equations in the case of a multi-electron system in three spatial dimensions interacting quantum electrodynamically with an arbitrary number of photonic modes. The resulting equations are similar in spirit to the equations of quantum electrodynamical density-functional theory~\cite{tokatly2013,ruggenthaler2014} in an optimized-effective potential scheme~\cite{pellegrini2015,flick2017c}, except that the orbitals used in that scheme are Kohn-Sham orbitals, while ours essentially are mean-field orbitals. We then derive the one-dimensional model used in the main text, as well as describe the parameters used in generating the data of Fig. 2 of the main text.

@@

\textcolor{blue}{As discussed in the main text, the variational energies, as well as the energies from numerical diagonalization, are compared against the energies calculated from perturbation theory. We provide additional details on the perturbative calculations here. What we calculate is the energy shift of a state $|a,0\rangle$, where $a$ is any matter state, and $0$ denotes the photonic vacuum, according to the perturbation Hamiltonian $\frac{q}{m}Ap + \frac{q^2}{2m}A^2$, where $p$ is the momentum operator for the matter defined in the main text, and $A$ is the vector potential operator of the one-dimensional cavity in the absence of the matter. In particular, $A(z) = \sum_n\sqrt{\frac{\hbar}{\epsilon_0\omega_n L}}\sin\left(\frac{n\pi z}{L} \right) \left(a_n+a^{\dagger}_n\right)$, with $L$ the length of the cavity and $\omega_n = \frac{n\pi c}{L}$. Upon inspection, it is clear that the $A^2$ term leads to an energy shift at first-order in perturbation theory, while the $A\cdot p$ term leads to an energy shift at second-order in perturbation theory. A straightforward application of perturbation theory leads to the following first-order energy shift from the $A^2$ term:
\begin{equation}
\delta E_a = \frac{q^2}{2m}\sum\limits_{n=1}^{N_p} \frac{\hbar}{\epsilon_0\omega_n L}\sin^2\left(\frac{n\pi z}{L}\right),
\end{equation}
where $z$ is the location of the matter in the cavity, and $N_p$ is the maximum photon number retained. Note that this value is dependent on the maximum photon number (although somewhat weakly) retained due to the logarithmically diverging nature of the sum, and so to compare variational, numerical, and perturbative calculations, the same number of modes are retained in each. A similarly straightforward application of perturbation theory leads to the energy shift coming from the $A\cdot p$ term being given by:
\begin{equation}
\delta E_a = \frac{q^2}{m^2}\sum\limits_{b\neq a}\sum\limits_{n=1}^{N_p} \frac{\hbar}{\epsilon_0\omega_n L} \frac{|p_{ba}|^2\sin^2\left(\frac{n\pi z}{L} \right)}{\hbar(\omega_a - \omega_b - \omega_n)},
\end{equation}
where $b$ denotes a matter state different from $a$, and $p_{ba} = \langle b|p|a\rangle$.}

\textcolor{blue}{In the main text, we also considered expectation values such as $\langle A \cdot p \rangle$ in the ground state within perturbation theory. Calculating the shift of the ground-state wavefunction at first-order in perturbation theory, we immediately find that
\begin{equation}
\langle Ap \rangle = \frac{2q}{m} \text{Re } \sum\limits_{b\neq a}\sum\limits_{n=1}^{N_p} \frac{1}{2\epsilon_0L}\frac{|p_{ba}|^2\sin^2\left(\frac{n\pi z}{L}\right)}{\omega_n(\omega_a-\omega_b-\omega_n)}.
\end{equation}}

\textcolor{blue}{It is also of interest to consider "field-only" observables such as the number of (virtual) photons in the ground state. Calculating the wavefunction of the ground state at first-order in perturbation theory, and calculating the probability of a photon being in the ground state, one finds that 
\begin{equation}
\langle N \rangle =\sum\limits_{n=1}^{N_p} \langle a^{\dagger}_{n,\text{bare}}a_{n,\text{bare}} \rangle = \frac{q^2}{m^2} \sum\limits_{b\neq a}\sum\limits_{n=1}^{N_p} \frac{1}{2\hbar\epsilon_0L}\frac{|p_{ba}|^2\sin^2\left(\frac{n\pi z}{L}\right)}{\omega_n(\omega_a-\omega_b-\omega_n)^2},
\end{equation}
where the "b" or "bare" subscript in the creation and annihilation operators denotes that these are creation and annihilation operators for photon modes in the cavity in the absence of interactions (e.g., the sine modes for the 1D cavity considered here). In Fig. S1, we compare the expectation value of the number of "bare" photons calculated  perturbatively, and with exact diagonalization with the variationally calculated number of "interacting" photons, $\sum\limits_{n=1}^{N_p} \langle a^{\dagger}_{n,\text{int}}a_{n,\text{int}} \rangle$ (e.g., the modes shown in Fig. 2b of the main text). The number of photons in the bare basis differs substantially from the number of photons in the interacting basis.}

\textcolor{blue}{This behavior does not reflect any inconsistency with the variational method, but is instead a feature which is to be expected. In particular, the physics of the ansatz of Eq. 4 of the main text is that a complicated ground state which has many virtual photons (see numerical diagonalization result) in it should be physically equivalent to a ground state with nearly no virtual photons (see variational result), but with different modes and frequencies. In other words, the reason the number of virtual photons differs is simply because they are being counted in the basis of "interacting modes" shown in Fig. 2b of the main text.   That said, our variational theory does not give a prescription to translate the interacting mode operators to the bare mode operators.}

\begin{figure}[t]
\includegraphics[width=6cm]{FigureS1.pdf}
\caption{\textcolor{blue}{\textbf{Number of virtual photons (bare and interacting) in the ground state calculated variationally, numerically, and through perturbation theory.} Parameters are the same as in Fig. 2 (top panel) of the main text.}}
\label{fig:ansatz}
\end{figure}

\textcolor{blue}{We note that in all cases, when calculating these quantities in the variational theory, these expressions are still relevant, with the replacement of the bare photon frequencies and mode functions replaced by the ones that result from Eq. 7 of the main text.}

@@

\textcolor{blue}{Regarding the numerical diagonalization, we numerically implement the Hamiltonian of Eq. (9) of the main text by building the matter operators ($H_0$ and $p$ as defined in the main text) in the basis of states corresponding to the tensor product of any matter state, and any photon state having four photons or less . Details regarding the ordering convention for the properly symmetrized multi-photon states are provided in Ref.~ \cite{flick2017} for example. Photon operators ($a$, $a^{\dagger}$) are constructed in this basis, and used to construct field operators such as $A$ in the basis of bare cavity modes (the usual sine modes). In particular, $A(z) = \sum_n\sqrt{\frac{\hbar}{\epsilon_0\omega_n L}}\sin\left(\frac{n\pi z}{L} \right) \left(a_n+a^{\dagger}_n\right)$, with $L$ the length of the cavity and $\omega_n = \frac{n\pi c}{L}$. Diagonalization is performed using standard sparse eigendecomposition routines (such as those implemented in MATLAB).}

>>>>

\maketitle
In this Supplement, we derive a self-consistent extension of the equations describing the ground state of a general quantum electrodynamical system with the effect of correlations included. We derive the equations in the case of a multi-electron system in three spatial dimensions interacting quantum electrodynamically with an arbitrary number of photonic modes. The resulting equations are similar in spirit to the equations of quantum electrodynamical density-functional theory~\cite{tokatly2013,ruggenthaler2014} in an optimized-effective potential scheme~\cite{pellegrini2015,flick2017c}, except that the orbitals used in that scheme are Kohn-Sham orbitals, while ours essentially are mean-field orbitals. We then derive the one-dimensional model used in the main text, as well as describe the parameters used in generating the data of Fig. 2 of the main text. Before doing this, we briefly describe the assumptions made in our framework, regarding our assumptions on gauge, as well as high-frequency free-space modes.

\subsection{Assumptions behind the Hamiltonian of Equations (1-3)}

\textcolor{blue}{In writing the Hamiltonian of the main text in the form of Equations (1-3), with the particular form of the field operators, we make a few assumptions. For one, that the interaction of matter with the quantized electromagnetic field is mediated solely by the vector potential. Moreover, we are assuming that the effect of free-space modes, which an emitter weakly interacts with, is taken into account by renormalization of the electron mass and charge \cite{milonni1976semiclassical}. We focus only on the interaction of the emitter with low-energy cavity modes where strong interaction is possible. For these cavity modes, a natural cutoff emerges based on the transparency of the cavity.}

\subsection{Generalized Coulomb gauge condition of Equation (7) of the main text}

\textcolor{blue}{In the main text, we arrive at the generalized Coulomb gauge for the modes. The generalized Coulomb gauge, $\nabla\cdot\epsilon(\mathbf{x})\mathbf{A}(\mathbf{x})=0$ for some dielectric permittivity $\epsilon(\mathbf{x})$, is the most common gauge for analyzing phenomena in the discipline of macroscopic QED, in which one seeks to analyze the interaction of charges with fields in dielectric media \cite{glauber1991quantum}. The reason a particular gauge was chosen was that in the Hamiltonian, we have already assumed that all interaction with the quantized modes is manifested only through the vector potential, and that there is no scalar potential contribution to the interaction: relaxing this assumption allows more general gauge freedom. We also remark that if we performed a frame transformation, for example to the length gauge, the energy would be the same as that of our ansatz in the velocity gauge, although the state would be markedly different, and would be a highly correlated state.}

@@

As discussed in the main text, the variational energies, as well as the energies from numerical diagonalization, are compared against the energies calculated from perturbation theory. We provide additional details on the perturbative calculations here. What we calculate is the energy shift of a state $|a,0\rangle$, where $a$ is any matter state, and $0$ denotes the photonic vacuum, according to the perturbation Hamiltonian $\frac{q}{m}Ap + \frac{q^2}{2m}A^2$, where $p$ is the momentum operator for the matter defined in the main text, and $A$ is the vector potential operator of the one-dimensional cavity in the absence of the matter. In particular, $A(z) = \sum_n\sqrt{\frac{\hbar}{\epsilon_0\omega_n L}}\sin\left(\frac{n\pi z}{L} \right) \left(a_n+a^{\dagger}_n\right)$, with $L$ the length of the cavity and $\omega_n = \frac{n\pi c}{L}$. Upon inspection, it is clear that the $A^2$ term leads to an energy shift at first-order in perturbation theory, while the $A\cdot p$ term leads to an energy shift at second-order in perturbation theory. A straightforward application of perturbation theory leads to the following first-order energy shift from the $A^2$ term:
\begin{equation}
\delta E_a = \frac{q^2}{2m}\sum\limits_{n=1}^{N_p} \frac{\hbar}{\epsilon_0\omega_n L}\sin^2\left(\frac{n\pi z}{L}\right),
\end{equation}
where $z$ is the location of the matter in the cavity, and $N_p$ is the maximum photon number retained. Note that this value is dependent on the maximum photon number (although somewhat weakly) retained due to the logarithmically diverging nature of the sum, and so to compare variational, numerical, and perturbative calculations, the same number of modes are retained in each. A similarly straightforward application of perturbation theory leads to the energy shift coming from the $A\cdot p$ term being given by:
\begin{equation}
\delta E_a = \frac{q^2}{m^2}\sum\limits_{b\neq a}\sum\limits_{n=1}^{N_p} \frac{\hbar}{\epsilon_0\omega_n L} \frac{|p_{ba}|^2\sin^2\left(\frac{n\pi z}{L} \right)}{\hbar(\omega_a - \omega_b - \omega_n)},
\end{equation}
where $b$ denotes a matter state different from $a$, and $p_{ba} = \langle b|p|a\rangle$.

In the main text, we also considered expectation values such as $\langle A \cdot p \rangle$ in the ground state within perturbation theory. Calculating the shift of the ground-state wavefunction at first-order in perturbation theory, we immediately find that
\begin{equation}
\langle Ap \rangle = \frac{2q}{m} \text{Re } \sum\limits_{b\neq a}\sum\limits_{n=1}^{N_p} \frac{1}{2\epsilon_0L}\frac{|p_{ba}|^2\sin^2\left(\frac{n\pi z}{L}\right)}{\omega_n(\omega_a-\omega_b-\omega_n)}.
\end{equation}

It is also of interest to consider "field-only" observables such as the number of (virtual) photons in the ground state. Calculating the wavefunction of the ground state at first-order in perturbation theory, and calculating the probability of a photon being in the ground state, one finds that 
\begin{equation}
\langle N \rangle =\sum\limits_{n=1}^{N_p} \langle a^{\dagger}_{n,\text{bare}}a_{n,\text{bare}} \rangle = \frac{q^2}{m^2} \sum\limits_{b\neq a}\sum\limits_{n=1}^{N_p} \frac{1}{2\hbar\epsilon_0L}\frac{|p_{ba}|^2\sin^2\left(\frac{n\pi z}{L}\right)}{\omega_n(\omega_a-\omega_b-\omega_n)^2},
\end{equation}
where the "b" or "bare" subscript in the creation and annihilation operators denotes that these are creation and annihilation operators for photon modes in the cavity in the absence of interactions (e.g., the sine modes for the 1D cavity considered here). In Fig. S1, we compare the expectation value of the number of "bare" photons calculated  perturbatively, and with exact diagonalization with the variationally calculated number of "interacting" photons, $\sum\limits_{n=1}^{N_p} \langle a^{\dagger}_{n,\text{int}}a_{n,\text{int}} \rangle$ (e.g., the modes shown in Fig. 2b of the main text). The number of photons in the bare basis differs substantially from the number of photons in the interacting basis.

This behavior does not reflect any inconsistency with the variational method, but is instead a feature which is to be expected. In particular, the physics of the ansatz of Eq. 4 of the main text is that a complicated ground state which has many virtual photons (see numerical diagonalization result) in it should be physically equivalent to a ground state with nearly no virtual photons (see variational result), but with different modes and frequencies. In other words, the reason the number of virtual photons differs is simply because they are being counted in the basis of "interacting modes" shown in Fig. 2b of the main text.   That said, our variational theory does not give a prescription to translate the interacting mode operators to the bare mode operators.

\begin{figure}[t]
\includegraphics[width=6cm]{FigureS1.pdf}
\caption{\textbf{Number of virtual photons (bare and interacting) in the ground state calculated variationally, numerically, and through perturbation theory.} Parameters are the same as in Fig. 2 (top panel) of the main text.}
\label{fig:ansatz}
\end{figure}

We note that in all cases, when calculating these quantities in the variational theory, these expressions are still relevant, with the replacement of the bare photon frequencies and mode functions replaced by the ones that result from Eq. 7 of the main text.

@@

Regarding the numerical diagonalization, we numerically implement the Hamiltonian of Eq. (9) of the main text by building the matter operators ($H_0$ and $p$ as defined in the main text) in the basis of states corresponding to the tensor product of any matter state, and any photon state having four photons or less . Details regarding the ordering convention for the properly symmetrized multi-photon states are provided in Ref.~ \cite{flick2017} for example. Photon operators ($a$, $a^{\dagger}$) are constructed in this basis, and used to construct field operators such as $A$ in the basis of bare cavity modes (the usual sine modes). In particular, $A(z) = \sum_n\sqrt{\frac{\hbar}{\epsilon_0\omega_n L}}\sin\left(\frac{n\pi z}{L} \right) \left(a_n+a^{\dagger}_n\right)$, with $L$ the length of the cavity and $\omega_n = \frac{n\pi c}{L}$. Diagonalization is performed using standard sparse eigendecomposition routines (such as those implemented in MATLAB).
